Adaption of the Chumbley Score to matching of bullet striation marks

### 2018-02-24 doc/paper.Rmd
additions to the introduction

## Motivation
\hh{XXX first sentence: Same source analyses are a major part of an Forensic Toolmark examiners job. XXX second sentence: describe standard practice: In current practice firearm examiners  make these comparisons visually under a comparison microscope and report XXX what is being reported: AFTE. XXX find: AFTE theory of identification as it relates to toolmarks  }
\hh{One of the main criticisms in this toolmark examination first raised by the National Research Council in 2009 [@NAS:2009] and later emphasized further by the President's Council of Advisors on Science and Technology [@pcast2016] are the amount of subjectivity in the assessment and the lack of error rates. }
\hh{XXX we have digitized images, e.g. profilometers or Confocal microscopy allow us to measure the surface [@vorburger2016]}

@@

Compairing pairs of toolmarks with the intention of matching it to a tool has been studied many times in the past. Extensive examples can be found in literature for tools and toolmark research ranging from screwdrivers \citep{manytoolmarks1, manytoolmarks2, chumbley} to groove pliers \citep{manytoolmarks2} to slip-joint pliers \citep{afte-chumbley} and many more. In comparison to this, same source matching of bullets to firearms has not been examined as prominently as that of toolmarks. Even less information is available on validity of methods and error rates associated with firearms examination. The National Academy of Sciences in its report in 2009 \citep{NAS:2009} discussed the need for determining error rates in methods proposed for firearms examination. As seen in the case of most forensic applications, the first step for same source matching and error rate determination involves identification of unique features that are characteristic of the object at hand. For the case of bullets and firearms, striation marks on the surface of the bullet are considered to be such markings that can be used in methods for same source matching. These marks are often a product of rifling and impurities and defects due to manufacturing in the barrel of the gun, which leads to engravings on the bullet surface \citep{afte-article1992}. In current practice, firearm examiners invariably make visual comparisons of bullet striae and use visual assesment tools to dignify bullets as being matches and non-matches. One way of accomplishing anykind of comparison between bullets is to do a comparison between surface marking of two or more bullet lands. Bullet Lands are considered to be areas between grooves made by the rifling action of the barrel and the markings on them are considered to be unique. The land engraved markings or sometimes termed as Bullet profiles \citep{aoas} are striation marks made on Bullet lands and often used for these land to land comparisons. Bullet Signatures is another word used in literature as seen in the work of \citet{chu2013} and \citet{aoas}. In our context bullet signatures refer to a processed version of the raw land engraved markings or profiles. The generation of bullet signatures involves first extraction of a bullet profile by taking the cross-sectional of the surface at a given height and then using loess fits to model the structure. The residuals of this fit are called signatures, which are considered to be noise free and a good reflection of the class charecteristics and unique features of a bullet. A more detailed version of the extraction technique of signatures is discussed by \citet{aoas}, where comprehensive details about the height at which profile is to be selected, removing curvature, smoothing, identifying groove locations are explained. Figure \ref{fig:rgl} shows us how the extracted signature from a bullet land lines up with the image of the land from which it was extracted. We can see also see in the figure how the depth and relative position of the striation markings seen in the image are interpreted as the signature.

>>>>

## Motivation
Same source analyses are a major part of an Forensic Toolmark examiners job. In current practice firearm examiners  make these comparisons visually under a comparison microscope and make one of the following four conclusions: identification, inconclusive, elimination or unsuitable for examination \citet{afte-website-2}. These conclusions are made on the basis of "unique surface contours" of the two toolmarks being in "sufficient agreement" \citep{afte-toolmarks1998}. AFTE describes the term "sufficient agreement" as the possibility of another tool producing the markings under comparison, as practically impossible \citep{afte-toolmarks1998} \citep{afte-website}. \gk{Although, they go on to concede that the interpretation of the identification depends on the examiner and is  therefore subjective in nature. }

One of the main criticisms in the toolmark examination first raised by the National Research Council in 2009 \citep{NAS:2009} and later emphasized further by the President's Council of Advisors on Science and Technology \citep{pcast2016} are the amount of subjectivity in the assessment and the lack of error rates.
Using a statistical approach removes both subjectivity from the assessment and allows us to provide error rates for both false positive and false negative identifications.

Digitized images, e.g. profilometers or confocal microscopy allow us to measure the surface \citep{vorburger2016} and can be used to conduct statistical analysis on the markings. Multiple toolmarks have been studied in the literature:  \citet{manytoolmarks1, chumbley} have studied screwdriver marks digitized using a profilometer,  \citet{manytoolmarks2} have investigated screwdriver, tongue and groove pliers marks that are generated using a confocal microscope,  while slip-joint pliers have been investigated by \citet{afte-chumbley} using digitized marks generated by a surface profilemeter. \citet{manytoolmarks1} used an algorithm which compares many small windows in the markings between two toolmarks and the maximum pearson correlation coefficient as the statistical index for concluding matches and non-matches. The Chumbley scoring method, first introduced by \citet{chumbley} uses a similar but much more extenstive framework which includes a non-parametric statistical test. This was later improved by Hadler and Morris \citep{hadler}. In this paper, we are investigating the applicability of the Chumbley scoring method by \citet{hadler} to assessing striation marks on bullets for same-source identification.

Striation marks on bullets are made by impurities in the barrel. As the bullet travels through the barrel, these imperfections leave "scratches" on the bullet surface. Typically, only striation marks in the land engraved areas (LEAs) are considered \citet{afte-article1992}. Bullet Lands are depressed areas between bullet grooves made by the rifling action of the barrel. Compared to toolmarks made by screwdrivers striation marks on bullets are typically much smaller, both in length and in width. Bullets also have a curved cross-sectional topography.

Bullet matching methods depend on the attribute being used for making comparisons. For single feature based matching \citet{chu2013} used an automatic method for counting consequtive matching striae (CMS). They reported zero false positive rate but do not discuss about the false negative rates for land to land matches. \citet{ma2004} and \citet{vorburger2011} discuss about CCF (cross-correlation function) and its discriminating power and application but do not talk about matches and non-matches. \citet{aoas} use multiple attributes like CCF, CMS and compare every land against every other land of the Hamby dataset \citet{hamby}. This also includes matching of all the unknown bullet lands against the known bullet lands. They report for 10,731 comparisons, 18 false negatives and 9 false negatives.

The Chumbley score  provides us with another approach in the same-source assessment of bullet striation marks.
\citet{chumbley}, in their paper, compare two toolmarks with the intention of determining if it comes from the same tool. The data for their study was obtained from 50 sequentially manufactured screwdriver tips. For every 100 comparisons \citet{chumbley} reported 2 false negatives and 9 false positives. The improvement proposed by \citet{hadler} on the other hand gave 0 false  negatives and 3 false poisitives.

\gk{\citep{afte-website} quotes sufficient agreement as "agreement of individual characteristics is of a quantity and quality that the likelihood another tool could have made the mark is so remote as to be considered a practical impossibility"}.} 
\gk{could not find the pdf for AFTE theory of identification as it relates to toolmarks pdf. I believe it is not available for free. But from the AFTE Journal the abstract was available which mentions the term " sufficient agreement" as quoted above. Currently the website seems like the most reliable way of quoting AFTE, i have checked other sources in both occasions where the website has been quoated above, like an FBI paper available at the FBI website and some non published but available pdf files in the AFTE website that have used these lines and quoted the AFTE journal. But since I cannot verify it in the Journal I have quoated the website.}
\gk{faden et al have not cited any paper that discusses the algorithm, i assume it was developed for the paper, the discussion of which is not as extensive as chumbley's. Its a variation of the chumbley method but from my understanding it is just the optimization step. and this paper came out much before the chumbley score.}

<!-- #### -->

\hh{XXX first sentence: Same source analyses are a major part of an Forensic Toolmark examiners job. XXX second sentence: describe standard practice: In current practice firearm examiners  make these comparisons visually under a comparison microscope and report XXX what is being reported: AFTE. XXX find: AFTE theory of identification as it relates to toolmarks  }
\hh{One of the main criticisms in this toolmark examination first raised by the National Research Council in 2009 [@NAS:2009] and later emphasized further by the President's Council of Advisors on Science and Technology [@pcast2016] are the amount of subjectivity in the assessment and the lack of error rates. }
\hh{XXX we have digitized images, e.g. profilometers or Confocal microscopy allow us to measure the surface [@vorburger2016]}

@@

Compairing pairs of toolmarks with the intention of matching it to a tool has been studied many times in the past. Extensive examples can be found in literature for tools and toolmark research ranging from screwdrivers \citep{manytoolmarks1, manytoolmarks2, chumbley} to groove pliers \citep{manytoolmarks2} to slip-joint pliers \citep{afte-chumbley} and many more. In comparison to this, same source matching of bullets to firearms has not been examined as prominently as that of toolmarks. Even less information is available on validity of methods and error rates associated with firearms examination. The National Academy of Sciences in its report in 2009 \citep{NAS:2009} discussed the need for determining error rates in methods proposed for firearms examination. As seen in the case of most forensic applications, the first step for same source matching and error rate determination involves identification of unique features that are characteristic of the object at hand. For the case of bullets and firearms, striation marks on the surface of the bullet are considered to be such markings that can be used in methods for same source matching. These marks are often a product of rifling and impurities and defects due to manufacturing in the barrel of the gun, which leads to engravings on the bullet surface \citep{afte-article1992}. In current practice, firearm examiners invariably make visual comparisons of bullet striae and use visual assesment tools to dignify bullets as being matches and non-matches. One way of accomplishing anykind of comparison between bullets is to do a comparison between surface marking of two or more bullet lands. Bullet Lands are considered to be areas between grooves made by the rifling action of the barrel and the markings on them are considered to be unique. The land engraved markings or sometimes termed as Bullet profiles \citep{aoas,ma2004} are striation marks made on Bullet lands and often used for these land to land comparisons. Bullet Signatures is another word used in literature as seen in the work of \citet{chu2013} and \citet{aoas}. In our context bullet signatures refer to a processed version of the raw land engraved markings or profiles. The generation of bullet signatures involves first extraction of a bullet profile by taking the cross-sectional of the surface at a given height and then using loess fits to model the structure. The residuals of this fit are called signatures, which are considered to be noise free and a good reflection of the class charecteristics and unique features of a bullet. A more detailed version of the extraction technique of signatures is discussed by \citet{aoas}, where comprehensive details about the height at which profile is to be selected, removing curvature, smoothing, identifying groove locations are explained. Figure \ref{fig:rgl} shows us how the extracted signature from a bullet land lines up with the image of the land from which it was extracted. We can see also see in the figure how the depth and relative position of the striation markings seen in the image are interpreted as the signature.
